---
title: "Data clean-up | 18S V7 | CAW-25-03"
author: "Olivier Laroche"
date: "January, 2025"
output: 
  distill::distill_article:
    toc: true
    theme: night
    highlight: pygments
    center: true
code_folding: true
---


### Loading libraries

```{r Loading libraries, message=FALSE, warning=FALSE, code_folding=TRUE, include=FALSE}
library(biohelper)
library(ggplot2)
library(dplyr)
library(data.table)
library(Biostrings)
library(janitor)
library(ggthemes)
library(DT)
library(tibble)
library(tidyverse)
library(UpSetR)
library(vegan)
library(plotly)
library(crosstalk)
library(ranacapa)
library(plyr)
library(glue)
library(theseus)
library(microViz)
```

### Setting up the environment
```{r}
path_metabarcoding_data = "../../../Metabarcoding_data/18S_V7/DADA2_results/"
path_metadata = "../../../Metabarcoding_data/short_amplicon_metadata_kelp.csv"
```


### Loading the ps object and adding metadata

```{r Loading files, message=FALSE, warning=FALSE, code_folding=TRUE}
ps = readRDS(paste0(path_metabarcoding_data,"ps_run.rds")) %>% microViz::phyloseq_validate()
ps

# !!! Add sequench blanks
metadata =  fread(path_metadata) %>% clean_names() %>% remove_empty(c("rows", "cols")) %>% 
  dplyr::filter(genes == "18sV7") %>% 
  dplyr::filter(sequench_id %in% sample_names(ps))

sample_names(ps)[sample_names(ps) %ni% metadata$sequench_id]

if(all(metadata$sequench_id %in% sample_names(ps))){
    cat("\nAll metadata samples are in the ASV table\n")
}else{
    cat("\nMissing samples in the ASV table?\n")
    cat(metadata$sequench_id[which(metadata$sequench_id %ni% sample_names(ps))])
}

if(all(sample_names(ps) %in% metadata$sequench_id)){
  cat("\nAll samples in the ASV table are in the metadata\n")
  metadata = metadata[match(sample_names(ps),metadata$sequench_id),]
} else{
    cat("\nMissing samples in the metadata table?\n")
    cat(sample_names(ps)[which(sample_names(ps) %ni% metadata$sequench_id)])
}

cat("\nAdding metadata to the ps object\n")
all(sample_names(ps) == metadata$sequench_id)
ps = merge_phyloseq(ps,sample_data(metadata) %>% magrittr::set_rownames(metadata$sequench_id))
ps@sam_data$original_sample_id = ps@sam_data$sample_id
ps@sam_data$sample_id = sample_names(ps)

ps@tax_table@.Data = ps@tax_table@.Data %>% as.data.frame() %>% dplyr::mutate("asv" = taxa_names(ps)) %>% as.matrix()
t = ps@tax_table@.Data %>% as.data.frame()
```


### Inspecting data and library size

```{r echo=FALSE, message=FALSE, warning=FALSE, code_folding=TRUE}
cat("\n\nData info\n")
ps
cat("\n\nMetadata info\n")
metadata %>% skimr::skim()

df <- ps %>% pstoveg_sample() 
if(is.null(df$sample_id)){
  df = df %>% rownames_to_column("sample_id")
}
df$LibrarySize <- sample_sums(ps)
df <- df[order(df$LibrarySize),]
df$Index <- seq(nrow(df))
p1=ggplot(data=df[df$amplicon_type == "sample",], aes(x=Index, y=LibrarySize, colour = amplicon_type, label = sample_id)) + geom_point() + theme_stata() + ggtitle("Sequencing depth per sample") +xlab("Sample index") + ylab("Library size") + scale_color_brewer(palette="Paired",name = "Amplicon type")
p2=ggplot(data=df[df$amplicon_type != "sample",], aes(x=Index, y=LibrarySize, colour = amplicon_type, label = sample_id)) + geom_point() + theme_stata() + ggtitle("Sequencing depth per blank") +xlab("Sample index") + ylab("Library size") + scale_color_brewer(palette="Paired",name = "Amplicon type")

ggplotly(p1)
ggplotly(p2)
```

### Inspection of contamination
#### Looking at reads and richness in blanks
```{r echo=FALSE, message=FALSE, warning=FALSE}
#Subsetting floc blanks
ps_blanks = ps %>% subset_samples(amplicon_type != "sample") %>% phyloseq::filter_taxa(function(x) sum(x) > 0, TRUE)
ps_blanks
# 
# # Creating a table
Contamination = ps_blanks %>% sample_names() %>% as.data.frame(stringsAsFactors=F)
names(Contamination) = "sample_name"
# ## Samples sum
Contamination$Read_count = ps_blanks %>% sample_sums() 
Contamination$Blank_type = ps_blanks@sam_data$amplicon_type
# 
# ## Diversity
R = estimate_richness(ps_blanks, measures="Observed")
Contamination$Richness = R$Observed
# 
# ## Ordering the table
Contamination = Contamination %>% dplyr::arrange(Richness)
# 
Contamination %>% datatable(class = 'cell-border stripe', caption = "Blanks and their read count and richness", options = list(pageLength = 100))
# 
dir.create(paste0(path_metabarcoding_data,"Contamination"))
write.table(Contamination,paste0(path_metabarcoding_data,"Contamination/Contamination.txt"), quote = F, sep = "\t", row.names = F)
```

#### Proportion of blank ASVs only in blanks

```{r echo=FALSE, message=FALSE, warning=FALSE, collapse=T}
ASVs_in_Blanks = taxa_names(ps_blanks)
ASVs_in_Blanks_only = ASVs_in_Blanks[which(ASVs_in_Blanks %ni% (ps %>%subset_samples(amplicon_type == "sample") %>% phyloseq::filter_taxa(function(x) sum(x) > 0, TRUE) %>% taxa_names()) )]

paste0("Percent of ASVs in blanks only found in blanks: ",round(length(ASVs_in_Blanks_only)/length(ASVs_in_Blanks)*100,2)," %")
```

#### Looking at the importance of ASVs found in contamination

```{r echo=FALSE, message=FALSE, warning=FALSE, collapse=T, fig.height=8, fig.width=10}
ps_samples_RA = ps %>% subset_samples(amplicon_type == "sample") %>% phyloseq::filter_taxa(function(x) sum(x) > 0, TRUE) %>% microbiome::transform(transform = "compositional")
ps_samples_RA_contam = ps_samples_RA %>% subset_taxa(asv %in% ASVs_in_Blanks)

tb = round(taxa_sums(ps_samples_RA_contam) / nsamples(ps_samples_RA_contam) *100,2) %>% as.data.frame() %>% setNames("Mean relative abundance per taxa (%)") %>% dplyr::arrange("Mean relative abundance per taxa (%)") 
tax = ps@tax_table@.Data %>% as.data.frame()
tb$taxo = ""

for (i in 1:nrow(tb)) {
  tb$taxo[i] = tax[tax$asv == rownames(tb)[i],] %>% unite('taxo', kingdom:genus, sep = ": ")
}

round(sample_sums(ps_samples_RA_contam) *100,2) %>% as.data.frame() %>% setNames("Mean relative abundance per sample (%)") %>% dplyr::arrange(desc(`Mean relative abundance per sample (%)`)) %>% datatable(class = 'cell-border stripe', caption = "Mean relative abundance of taxa found in blanks in samples", options = list(pageLength = 10))

df_ps_samples_RA_contam = ps_samples_RA_contam %>% psmelt()

fig2 <- plot_ly(df_ps_samples_RA_contam, x = ~sample_id, y = ~Abundance*100, color = ~family, type = 'bar') %>% layout(yaxis = list(title = 'Relative abundance (%)'), barmode = 'stack') %>% layout(legend=list(title=list(text='<b> Family </b>')))
fig2

ASVs_in_Blanks_only2 = speedyseq::psmelt(ps_samples_RA_contam)
ASVs_in_Blanks_only2 = ASVs_in_Blanks_only2 %>% dplyr::group_by(kingdom, phylum, class, order, family, genus, asv) %>% dplyr::summarise(Mean_RA = round(mean(Abundance)*100,2), SD = sd(Abundance)*100 ) %>% dplyr::arrange(desc(Mean_RA))

write.table(ASVs_in_Blanks_only2, paste0(path_metabarcoding_data,"Contamination/Taxo_blank_ASVs_in_samples.txt"), quote = F, row.names = F, sep = "\t")
```

#### Viewing blanks composition
```{r echo=FALSE, message=FALSE, warning=FALSE}
ps_blanks = ps %>% subset_samples(amplicon_type %ni% c("sample","Sample")) %>% phyloseq::filter_taxa(function(x) sum(x) > 0, TRUE) %>% microbiome::transform(transform = "compositional")
ps_blanks@refseq =NULL

ps_merged = ps %>% subset_samples(amplicon_type %in% c("sample","Sample")) %>% merge_samples("amplicon_type") %>% phyloseq::filter_taxa(function(x) sum(x) > 0, TRUE) %>% microbiome::transform(transform = "compositional")

ps_blanks = merge_phyloseq(ps_blanks,ps_merged) %>% microbiome::transform(transform = "compositional")

pblank = taxo_bar_plot(ps_obj = ps_blanks, rank1 = "phylum", rank2 = "family", label = "amplicon_type")
ggplotly(pblank)
```


#### Effect of removing blanks following different methodologies

```{r echo=FALSE}
cat("\nRemoving all taxa found in blanks\n")
ps_trimmed = ps %>% biohelper::ps_decon(method = "complete_asv_removal")

cat("\nRemoving reads from taxa found in blanks using max_v\n")
ps_trimmed = ps %>% biohelper::ps_decon(method = "max_v")

ps@sam_data$grp = paste0(ps@sam_data$extraction_blank_number, "_",ps@sam_data$pcr_batches)
cat("\nRemoving reads from taxa found in blanks using microDecon\n")
ps_trimmed = ps %>% biohelper::ps_decon(method = "microDecon", group = "grp")
```



### Looking at eukaryotic and non-eukaryotic taxa

```{r echo=FALSE, fig.height=5, fig.width=10, collapse=TRUE}
paste0("Number of reads from Eukaryotes: ",ps_trimmed %>% subset_taxa(superkingdom == "Eukaryota" | family %ni% "Hominidae") %>% sample_sums() %>% sum())
paste0("Number of reads from non-eukaryotes: ",ps_trimmed %>% subset_taxa(superkingdom != "Eukaryota" | family %in% "Hominidae")%>% sample_sums() %>% sum())

Non_euk_ASVs = ps_trimmed %>% subset_taxa(superkingdom != "Eukaryota" | family %in% "Hominidae") %>% taxa_names()

ps_trimmed_non_euk = ps_trimmed %>% transform_sample_counts(function(x) 100 * x/sum(x)) %>% subset_taxa(superkingdom!="Eukaryota"| family %in% "Hominidae") 

ps_trimmed_non_euk@tax_table@.Data = ps_trimmed_non_euk@tax_table@.Data %>% as.data.frame() %>% replace(is.na(.), "NA") %>% as.matrix()
tax2 = ps_trimmed_non_euk@tax_table@.Data %>% as.data.frame() 

plot_bar(ps_trimmed_non_euk, 'sample_id', fill = "phylum", title = "Non-eukaryotes reads") + ylab("Reads (%)") + theme_calc() + theme(axis.text.x = element_text(angle = 90, size=4)) + xlab("Sample unique ID")

# Looking at percent of ASVs belonging to non-Eukaryota
Non_euk_ASVs_nb = ps_trimmed %>% subset_taxa(superkingdom != "Eukaryota" | family %in% "Hominidae") %>% phyloseq::filter_taxa(function(x) sum(x) > 0, TRUE) %>% taxa_names() %>% length()
Euk_ASVs_nb = ps_trimmed %>% subset_taxa(superkingdom == "Eukaryota" | family %ni% "Hominidae") %>% phyloseq::filter_taxa(function(x) sum(x) > 0, TRUE) %>% taxa_names() %>% length()
paste0("Percent of non-eukaryotes ASVs: ",round(Non_euk_ASVs_nb/Euk_ASVs_nb *100,2)," %")

# Looking at percent of reads belonging to non-Eukaryota
Non_euk_reads_nb = ps_trimmed %>% subset_taxa(superkingdom != "Eukaryota" | family %in% "Hominidae") %>% phyloseq::filter_taxa(function(x) sum(x) > 0, TRUE) %>% sample_sums() %>% sum()
Euk_reads_nb = ps_trimmed %>% subset_taxa(superkingdom == "Eukaryota" | family %ni% "Hominidae") %>% phyloseq::filter_taxa(function(x) sum(x) > 0, TRUE) %>% sample_sums() %>% sum()
paste0("Percent of non-eukaryotes reads: ",round(Non_euk_reads_nb/Euk_reads_nb *100,5)," %")
```

#### Removing non-eukaryotic ASVs (including homonidae)

```{r echo=FALSE, message=FALSE, warning=FALSE, collapse=T}
ps_trimmed_euk = ps_trimmed %>% subset_taxa(superkingdom=="Eukaryota" & family %ni% "Hominidae" & class %ni% "Aves") 
# # Looking at read counts
# cat("\nBefore data\n")
# ps_trimmed
# paste0("Total number of reads: ",ps_trimmed %>% sample_sums() %>% sum())
# cat("\nAfter data\n")
# ps_trimmed_euk
# paste0("Total number of reads: ",ps_trimmed_euk %>% sample_sums() %>% sum())

# Adding reads lost for sediment samples (processed in another script)
before = ps_trimmed %>% sample_sums() %>% sum()
after = ps_trimmed_euk %>% sample_sums() %>% sum()
before_r = ps_trimmed %>% taxa_names() %>% length()
after_r = ps_trimmed_euk %>% taxa_names() %>% length()

cat("\nTotal number of lost ASVs and reads after non-eukaryote ASVs removal\n")
paste0("Lost reads: ", before-after)
paste0("Lost ASVs: ", before_r-after_r)

cat("\nPercent of lost ASVs and reads after non-eukaryote ASVs removal\n")
paste0("Lost reads: ",round(100-(after/before * 100),2)," %")
paste0("Lost ASVs: ",round(100-(after_r/before_r * 100),2)," %")
```

### Looking at number of rare ASVs

```{r echo=FALSE, fig.height=6, fig.width=10, message=FALSE, warning=FALSE, collapse=TRUE}
TS = sort(taxa_sums(ps_trimmed_euk)) %>% as.data.frame() %>% rownames_to_column("ASV")

p1= ggplot(TS, aes(.)) + geom_histogram() + theme_stata() + scale_x_sqrt(limits= c(0,10000), breaks = c(1,5,10,500,2500,5000,7500,10000)) + scale_y_sqrt(breaks = c(100, 1000, 5000,10000,25000,50000)) + xlab("Reads/ASV") + ylab("Number of ASVs") + ggtitle("ASVs distribution")

cat("\nProportions of rare ASVs in samples\n")

paste0("Singletons: ",taxa_sums(ps_trimmed_euk)[taxa_sums(ps_trimmed_euk)  == 1] %>% length(),": ",round(taxa_sums(ps_trimmed_euk)[taxa_sums(ps_trimmed_euk)  == 1] %>% length()/taxa_sums(ps_trimmed_euk)%>% length()*100,2)," %")

paste0("Doubletons: ",taxa_sums(ps_trimmed_euk)[taxa_sums(ps_trimmed_euk)  == 2] %>% length(),": ",round(taxa_sums(ps_trimmed_euk)[taxa_sums(ps_trimmed_euk)  == 2] %>% length()/taxa_sums(ps_trimmed_euk)%>% length()*100,2)," %")

paste0("Less than 10 reads: ",taxa_sums(ps_trimmed_euk)[taxa_sums(ps_trimmed_euk)  <10] %>% length(),": ",round(taxa_sums(ps_trimmed_euk)[taxa_sums(ps_trimmed_euk)  <10] %>% length()/taxa_sums(ps_trimmed_euk)%>% length()*100,2)," %")

p1
```

### Effect of removing rare ASVs 

```{r, echo=FALSE, message=FALSE, warning=FALSE, collapse=TRUE}
# Filtering taxa -------------------------------------------------
ps_trimmed_euk_min = phyloseq::filter_taxa(ps_trimmed_euk %>% phyloseq_validate(), function(x) sum(x > 10) > (1), TRUE)

# Adding reads lost for sediment samples (processed in another script)
before = ps_trimmed_euk %>% sample_sums() %>% sum()
after = ps_trimmed_euk_min %>% sample_sums() %>% sum()

cat("\nPercent of lost reads after rare ASVs removal\n")
paste0("Lost reads: ",round(100-(after/before * 100),2)," %")
cat("\nPercent of lost ASVs after rare ASVs removal\n")
paste0("Lost ASVs: ",round(100-(ntaxa(ps_trimmed_euk_min)/ntaxa(ps_trimmed_euk) * 100),2)," %")
```

### Looking at sequencing depth per sample with rarefaction curves

```{r , message=FALSE, warning=FALSE, collapse=TRUE, include=FALSE}
ps_trimmed_euk_min = ps_trimmed_euk # Keeping rare taxa for now
ps_trimmed_euk_min <- prune_samples(sample_sums(ps_trimmed_euk_min)>100, ps_trimmed_euk_min)

dir.create(paste0(path_metabarcoding_data,"/Cleaned_data"))
dir.create(paste0(path_metabarcoding_data,"/Cleaned_data/Sequencing_depth"))

s = sample_sums(ps_trimmed_euk_min) %>% sort()
x = split(s, ceiling(seq_along(s)/10))
xx <- plyr::ldply(x, function(x){
    plyr::rename(ldply(x, function(z) {data.frame(read_count=z)}),c(".id" = ".id2"))
})
xx =xx %>% dplyr::group_by(.id) %>% dplyr::mutate(.id3 = max(read_count))
xx = xx[match(sample_names(ps_trimmed_euk_min),xx$.id2),]
ps_trimmed_euk_min@sam_data$read_count_gr = paste0("Group_",xx$.id," (<=",xx$.id3,")")
ps_trimmed_euk_min@sam_data$read_count_gr = factor(ps_trimmed_euk_min@sam_data$read_count_gr, levels = gtools::mixedsort(ps_trimmed_euk_min@sam_data$read_count_gr %>% unique()))
p <- ranacapa::ggrare(ps_trimmed_euk_min, step = 1000, color = "read_count_gr",  se = FALSE)
p <- p + facet_wrap(~read_count_gr,scales = "free_x") + theme_bw() + ylab("ASV richness") + xlab("Sequencing depth") + guides(color=guide_legend(title="Sequencing depth")) + theme(axis.text.x = element_text(angle = 45))
ggsave(plot = p, filename = "rarefaction_curves.pdf", path = paste0(path_metabarcoding_data,"/Cleaned_data/Sequencing_depth/"), device = "pdf", width = 12, height = 8)
```

```{r echo=FALSE, fig.height= 8, fig.width=16}
ggplotly(p)
```


### Looking at read count per sample and removing those with less than X reads.

```{r echo=FALSE, message=FALSE, warning=FALSE}
df_read_count = ps_trimmed_euk_min %>% phyloseq::sample_sums() %>% sort() %>% as.data.frame(col.names = "Reads")
colnames(df_read_count) = "Read count"
df_read_count %>% datatable(class = 'cell-border stripe', caption = "Read count per sample after decontamination", options = list(pageLength = 10))

# Filtering samples -------------------------------------------------
ps_trimmed_euk_min_min = phyloseq::prune_samples(phyloseq::sample_sums(ps_trimmed_euk_min) >= 2000, ps_trimmed_euk_min)
```

### Saving results

```{r message=FALSE, warning=FALSE, code_folding=TRUE}
ps_trimmed_euk_min_min
saveRDS(ps_trimmed_euk_min_min,paste0(path_metabarcoding_data,"/Cleaned_data/ps_clean.rds"))

# # Txt tables ----------------------------------------------------
ASV_table = ps_trimmed_euk_min_min %>% pstoveg_otu() %>% t() %>% as.data.frame() %>% rownames_to_column('feature-id') 
write.table(ASV_table,paste0(path_metabarcoding_data,'/Cleaned_data/ASV_table.txt'), quote = F, row.names = F, sep = "\t")

Taxo_table = ps_trimmed_euk_min_min@tax_table@.Data %>% as.data.frame()
Taxo_table  = Taxo_table %>% rownames_to_column('feature-id')
write.table(Taxo_table,paste0(path_metabarcoding_data,"/Cleaned_data/taxonomy_table.txt"), quote = F, row.names = F, sep = "\t")

Metadata = ps_trimmed_euk_min_min@sam_data %>% data.frame()
Metadata  = Metadata %>% rownames_to_column('sample-id')
write.table(Metadata,paste0(path_metabarcoding_data,"/Cleaned_data/metadata_table.txt"), quote = F, row.names = F, sep = "\t")

writeXStringSet(ps_trimmed_euk_min_min@refseq,paste0(path_metabarcoding_data,"/Cleaned_data/ref_seq.fasta"))
```








