---
title: "Data clean-up | COI"
author: "Olivier Laroche"
date: "September, 2024"
output: 
  distill::distill_article:
    toc: true
    theme: night
    highlight: pygments
    center: true
code_folding: true
---

### Loading libraries

```{r Loading libraries, message=FALSE, warning=FALSE, code_folding=TRUE, include=FALSE}
library(biohelper)
library(ggplot2)
library(dplyr)
library(data.table)
library(Biostrings)
library(janitor)
library(ggthemes)
library(DT)
library(tibble)
library(tidyverse)
library(UpSetR)
library(vegan)
library(plotly)
library(crosstalk)
library(ranacapa)
library(plyr)
library(glue)
library(theseus)
library(microViz)
```

### Setting up the environment
```{r}
path_metabarcoding_data = "../../../Data/metabarcoding/COI/DADA2_results/"
path_metadata = "../../../Data/metadata/metadata.xlsx"
excel_sheet = "meta_metabarcoding" # if applicable
```

### Loading the ps object and adding metadata

```{r Loading files, message=FALSE, warning=FALSE, code_folding=TRUE}
ps = readRDS(paste0(path_metabarcoding_data,"ps_run.rds")) %>% microViz::phyloseq_validate()
ps
#ps@refseq %>% ape::as.DNAbin()

metadata =  readxl::read_excel(path_metadata,sheet = excel_sheet)%>%clean_names()%>%remove_empty(c("rows", "cols")) %>% 
  dplyr::filter(target_gene == "COI")

### Extra code to complete metadata
coordinate_func = function(obj){
  if(any(class(obj)%in%"phyloseq")){
    m_temp = obj %>% 
      pstoveg_sample() %>% 
      sf::st_as_sf(coords = c("easting_utm_11n_wgs84", "northing_utm_11n_wgs84"),
           crs = "+proj=utm +zone=11",
           na.fail = F) %>%
      sf::st_transform(crs = "+proj=longlat +datum=WGS84") %>%
      as.data.frame() %>% 
      dplyr::mutate(lon = unlist(map(geometry,1)),lat = unlist(map(geometry,2))) 
    m_temp$station_loc <- sf::st_geometry(sf::st_as_sf(m_temp,coords = c("lon","lat"),na.fail = F)) # Same as geometry so probably not needed
    #m_temp$geometry=NULL
    obj@sam_data = sample_data(m_temp)    
  }else{
    m_temp = obj %>% 
      sf::st_as_sf(coords = c("easting_utm_11n_wgs84", "northing_utm_11n_wgs84"),
           crs = "+proj=utm +zone=11",
           na.fail = F) %>%
      sf::st_transform(crs = "+proj=longlat +datum=WGS84") %>%
      as.data.frame() %>% 
      dplyr::mutate(lon = unlist(map(geometry,1)),lat = unlist(map(geometry,2))) 
    #m_temp$station_loc <- sf::st_geometry(sf::st_as_sf(m_temp,coords = c("lon","lat"),na.fail = F)) # Same as geometry so probably not needed
    m_temp$geometry=NULL
    m_temp$time_on_seafloor_utc = as.character(m_temp$time_on_seafloor_utc)
    obj = m_temp
  }
  return(obj)
}
metadata_extra = readxl::read_xlsx(paste0(path_metadata),sheet = "extra_info_eDNA", col_names = T) %>% 
  janitor::clean_names() %>% 
  dplyr::mutate(station_id = gsub("-","_",station_id))
metadata = left_join(metadata, metadata_extra, by = join_by(station_id, multicore_id)) %>% 
  coordinate_func()

head(sample_names(ps))
head(metadata$sample_id)

ps = ps %>% 
  subset_samples(SAMPLE %in% metadata$sample_id) %>% 
  phyloseq::filter_taxa(function(x) sum(x) > 0, TRUE)
ps

# Loading the Read count files from the bioinfo pre-processing
rd = fread(paste0(path_metabarcoding_data,"Read_counts.csv"))
colnames(rd)[1] = "sample_id"
rd$percent_retained_reads = round(rd$nonchim / rd$input * 100,2)
rd <- rd[gtools::mixedorder(rd$sample_id, decreasing = T), ]
write_csv(x = rd,file = "../Data_clean-up/rd_CAW-24-18_COI.csv")


# Adding the read count data to the metadata
metadata = left_join(metadata, rd, by = "sample_id")

if(all(metadata$sample_id %in% sample_names(ps))){
    cat("\nAll metadata samples are in the ASV table\n")
}else{
    cat("\nMissing samples in the ASV table?\n")
    cat(metadata$sample_id[which(metadata$sample_id %ni% sample_names(ps))])
}

if(all(sample_names(ps) %in% metadata$sample_id)){
  cat("\nAll samples in the ASV table are in the metadata\n")
  metadata = metadata[match(sample_names(ps),metadata$sample_id),]
} else{
    cat("\nMissing samples in the metadata table?\n")
    cat(sample_names(ps)[which(sample_names(ps) %ni% metadata$sample_id)])
}

cat("\nAdding metadata to the ps object\n")
all(sample_names(ps) == metadata$sample_id)
ps = merge_phyloseq(ps,sample_data(metadata) %>% magrittr::set_rownames(metadata$sample_id))
ps@sam_data$sample_id = sample_names(ps)

# Adding ASV names in taxa table
ps@tax_table@.Data = ps@tax_table@.Data %>% as.data.frame() %>% dplyr::mutate("asv" = taxa_names(ps)) %>% as.matrix()

# Creating Appendix 2
write_csv(x = metadata, file = paste0(path_metabarcoding_data,"Appendix2.csv"))
```


### Inspecting data and library size

```{r echo=FALSE, message=FALSE, warning=FALSE, code_folding=TRUE}
cat("\n\nData info\n")
ps
cat("\n\nMetadata info\n")
metadata %>% skimr::skim()

df <- ps %>% pstoveg_sample() 
if(is.null(df$sample_id)){
  df = df %>% rownames_to_column("sample_id")
}
df$LibrarySize <- sample_sums(ps)
df <- df[order(df$LibrarySize),]
df$Index <- seq(nrow(df))
p1=ggplot(data=df[df$amplicon_type == "sample",], aes(x=Index, y=LibrarySize, colour = amplicon_type, label = sample_id)) + geom_point() + theme_stata() + ggtitle("Sequencing depth per sample") +xlab("Sample index") + ylab("Library size") + scale_color_brewer(palette="Paired",name = "Amplicon type")
p2=ggplot(data=df[df$amplicon_type != "sample",], aes(x=Index, y=LibrarySize, colour = amplicon_type, label = sample_id)) + geom_point() + theme_stata() + ggtitle("Sequencing depth per blank") +xlab("Sample index") + ylab("Library size") + scale_color_brewer(palette="Paired",name = "Amplicon type")

ggplotly(p1)
ggplotly(p2)
```


### Inspection of contamination
#### Looking at reads and richness in blanks

```{r echo=FALSE, message=FALSE, warning=FALSE}
# #Subsetting floc blanks
# ps_blanks = ps %>% subset_samples(amplicon_type != "sample") %>% phyloseq::filter_taxa(function(x) sum(x) > 0, TRUE)
# ps_blanks
# # 
# # # Creating a table
# Contamination = ps_blanks %>% sample_names() %>% as.data.frame(stringsAsFactors=F)
# names(Contamination) = "sample_name"
# # ## Samples sum
# Contamination$Read_count = ps_blanks %>% sample_sums() 
# Contamination$Blank_type = ps_blanks@sam_data$amplicon_type
# # 
# # ## Diversity
# R = estimate_richness(ps_blanks, measures="Observed")
# Contamination$Richness = R$Observed
# # 
# # ## Ordering the table
# Contamination = Contamination %>% dplyr::arrange(Richness)
# # 
# Contamination %>% datatable(class = 'cell-border stripe', caption = "Blanks and their read count and richness", options = list(pageLength = 100))
# # 
# dir.create(paste0(path_metabarcoding_data,"Contamination"))
# write.table(Contamination,paste0(path_metabarcoding_data,"Contamination/Contamination.txt"), quote = F, sep = "\t", row.names = F)
```

#### Proportion of blank ASVs only in blanks

```{r echo=FALSE, message=FALSE, warning=FALSE, collapse=T}
# ASVs_in_Blanks = taxa_names(ps_blanks)
# ASVs_in_Blanks_only = ASVs_in_Blanks[which(ASVs_in_Blanks %ni% (ps %>%subset_samples(amplicon_type == "sample") %>% phyloseq::filter_taxa(function(x) sum(x) > 0, TRUE) %>% taxa_names()) )]
# 
# paste0("Percent of ASVs in blanks only found in blanks: ",round(length(ASVs_in_Blanks_only)/length(ASVs_in_Blanks)*100,2)," %")
```

#### Looking at the importance of ASVs found in contamination

```{r echo=FALSE, message=FALSE, warning=FALSE, collapse=T, fig.height=8, fig.width=10}
# ps_samples_RA = ps %>% subset_samples(amplicon_type == "sample") %>% phyloseq::filter_taxa(function(x) sum(x) > 0, TRUE) %>% microbiome::transform(transform = "compositional")
# ps_samples_RA_contam = ps_samples_RA %>% subset_taxa(asv %in% ASVs_in_Blanks)
# 
# tb = round(taxa_sums(ps_samples_RA_contam) / nsamples(ps_samples_RA_contam) *100,2) %>% as.data.frame() %>% setNames("Mean relative abundance per taxa (%)") %>% dplyr::arrange("Mean relative abundance per taxa (%)") 
# tax = ps@tax_table@.Data %>% as.data.frame()
# tb$taxo = ""
# 
# for (i in 1:nrow(tb)) {
#   tb$taxo[i] = tax[tax$asv == rownames(tb)[i],] %>% unite('taxo', Kingdom:Genus, sep = ": ")
# }
# 
# round(sample_sums(ps_samples_RA_contam) *100,2) %>% as.data.frame() %>% setNames("Mean relative abundance per sample (%)") %>% dplyr::arrange(desc(`Mean relative abundance per sample (%)`)) %>% datatable(class = 'cell-border stripe', caption = "Mean relative abundance of taxa found in blanks in samples", options = list(pageLength = 10))
# 
# df_ps_samples_RA_contam = ps_samples_RA_contam %>% psmelt()
# 
# fig2 <- plot_ly(df_ps_samples_RA_contam, x = ~sample_id, y = ~Abundance*100, color = ~Family, type = 'bar') %>% layout(yaxis = list(title = 'Relative abundance (%)'), barmode = 'stack') %>% layout(legend=list(title=list(text='<b> Family </b>')))
# #fig2
# 
# ASVs_in_Blanks_only2 = speedyseq::psmelt(ps_samples_RA_contam)
# ASVs_in_Blanks_only2 = ASVs_in_Blanks_only2 %>% dplyr::group_by(Kingdom, Phylum, Class, Order, Family, Genus, OTU) %>% dplyr::summarise(Mean_RA = round(mean(Abundance)*100,2), SD = sd(Abundance)*100 ) %>% dplyr::arrange(desc(Mean_RA))
# 
# write.table(ASVs_in_Blanks_only2, paste0(path_metabarcoding_data,"Contamination/Taxo_blank_ASVs_in_samples.txt"), quote = F, row.names = F, sep = "\t")
```

### Effect of removing blanks following different methodologies
Here, there was no contamination so just removing the blanks
```{r echo=FALSE}
# cat("\nRemoving all taxa found in blanks\n")
# ps_trimmed = ps %>% biohelper::ps_decon(method = "complete_asv_removal")
# 
# cat("\nRemoving reads from taxa found in blanks using max_v\n")
# ps_trimmed = ps %>% biohelper::ps_decon(method = "max_v")
# 
# cat("\nRemoving reads from taxa found in blanks using microDecon\n")
# ps_trimmed = ps %>% biohelper::ps_decon(method = "microDecon") #, group = "extraction_blanks"

ps_trimmed = ps %>% subset_samples(amplicon_type %in% c("sample","Sample")) %>% phyloseq::filter_taxa(function(x) sum(x) > 0, TRUE)
```


### Looking at eukaryotic and non-eukaryotic taxa

```{r echo=FALSE, fig.height=5, fig.width=10, message=FALSE, warning=FALSE, collapse=TRUE}
paste0("Number of reads from Eukaryotes: ",ps_trimmed %>% subset_taxa(superkingdom == "Eukaryota" | order %ni% c("Primates","Artiodactyla")) %>% sample_sums() %>% sum())
paste0("Number of reads from non-eukaryotes: ",ps_trimmed %>% subset_taxa(superkingdom != "Eukaryota" | order %in% c("Primates","Artiodactyla"))%>% sample_sums() %>% sum())

Non_euk_ASVs = ps_trimmed %>% subset_taxa(superkingdom != "Eukaryota" | order %in% c("Primates","Artiodactyla")) %>% taxa_names()

ps_trimmed_non_euk = ps_trimmed %>% transform_sample_counts(function(x) 100 * x/sum(x)) %>% subset_taxa(superkingdom!="Eukaryota"| order %in% c("Primates","Artiodactyla"))

ps_trimmed_non_euk@tax_table@.Data = ps_trimmed_non_euk@tax_table@.Data %>% as.data.frame() %>% replace(is.na(.), "NA") %>% as.matrix()
tax2 = ps_trimmed_non_euk@tax_table@.Data %>% as.data.frame() 

plot_bar(ps_trimmed_non_euk, 'sample_id', fill = "phylum", title = "Non-eukaryotes reads") + ylab("Reads (%)")  + theme(axis.text.x = element_text(angle = 90, size=4)) + xlab("Sample unique ID")

# Looking at percent of ASVs belonging to non-Eukaryota
Non_euk_ASVs_nb = ps_trimmed %>% subset_taxa(superkingdom != "Eukaryota" | order %in% c("Primates","Artiodactyla")) %>% phyloseq::filter_taxa(function(x) sum(x) > 0, TRUE) %>% taxa_names() %>% length()
Euk_ASVs_nb = ps_trimmed %>% subset_taxa(superkingdom == "Eukaryota" | order %ni% c("Primates","Artiodactyla")) %>% phyloseq::filter_taxa(function(x) sum(x) > 0, TRUE) %>% taxa_names() %>% length()
paste0("Percent of non-eukaryotes ASVs: ",round(Non_euk_ASVs_nb/Euk_ASVs_nb *100,2)," %")

# Looking at percent of reads belonging to non-Eukaryota
Non_euk_reads_nb = ps_trimmed %>% subset_taxa(superkingdom != "Eukaryota" | order %in% c("Primates","Artiodactyla")) %>% phyloseq::filter_taxa(function(x) sum(x) > 0, TRUE) %>% sample_sums() %>% sum()
Euk_reads_nb = ps_trimmed %>% subset_taxa(superkingdom == "Eukaryota" | order %ni% c("Primates","Artiodactyla")) %>% phyloseq::filter_taxa(function(x) sum(x) > 0, TRUE) %>% sample_sums() %>% sum()
paste0("Percent of undesired or non-eukaryotes reads: ",round(Non_euk_reads_nb/Euk_reads_nb *100,5)," %")
```

#### Removing undesired and non-eukaryotic ASVs (including homonidae)

```{r echo=FALSE, message=FALSE, warning=FALSE, collapse=T}
ps_trimmed_euk = ps_trimmed %>% subset_taxa(superkingdom =="Eukaryota" & order %ni% c("Primates","Artiodactyla")) 
# # Looking at read counts
# cat("\nBefore data\n")
# ps_trimmed
# paste0("Total number of reads: ",ps_trimmed %>% sample_sums() %>% sum())
# cat("\nAfter data\n")
# ps_trimmed_euk
# paste0("Total number of reads: ",ps_trimmed_euk %>% sample_sums() %>% sum())

# Adding reads lost for sediment samples (processed in another script)
before = ps_trimmed %>% sample_sums() %>% sum()
after = ps_trimmed_euk %>% sample_sums() %>% sum()
before_r = ps_trimmed %>% taxa_names() %>% length()
after_r = ps_trimmed_euk %>% taxa_names() %>% length()

cat("\nTotal number of lost ASVs and reads after non-eukaryote ASVs removal\n")
paste0("Lost reads: ", before-after)
paste0("Lost ASVs: ", before_r-after_r)

cat("\nPercent of lost ASVs and reads after non-eukaryote ASVs removal\n")
paste0("Lost reads: ",round(100-(after/before * 100),2)," %")
paste0("Lost ASVs: ",round(100-(after_r/before_r * 100),2)," %")
```

### Looking at number of rare ASVs

```{r echo=FALSE, fig.height=6, fig.width=10, message=FALSE, warning=FALSE, collapse=TRUE}
TS = sort(taxa_sums(ps_trimmed_euk)) %>% as.data.frame() %>% rownames_to_column("ASV")

p1= ggplot(TS, aes(.)) + geom_histogram() + scale_x_sqrt(limits= c(0,10000), breaks = c(1,5,10,500,2500,5000,7500,10000)) + scale_y_sqrt(breaks = c(100, 1000, 5000,10000,25000,50000)) + xlab("Reads/ASV") + ylab("Number of ASVs") + ggtitle("ASVs distribution")

cat("\nProportions of rare ASVs in samples\n")

paste0("Singletons: ",taxa_sums(ps_trimmed_euk)[taxa_sums(ps_trimmed_euk)  == 1] %>% length(),": ",round(taxa_sums(ps_trimmed_euk)[taxa_sums(ps_trimmed_euk)  == 1] %>% length()/taxa_sums(ps_trimmed_euk)%>% length()*100,2)," %")

paste0("Doubletons: ",taxa_sums(ps_trimmed_euk)[taxa_sums(ps_trimmed_euk)  == 2] %>% length(),": ",round(taxa_sums(ps_trimmed_euk)[taxa_sums(ps_trimmed_euk)  == 2] %>% length()/taxa_sums(ps_trimmed_euk)%>% length()*100,2)," %")

paste0("Less than 10 reads: ",taxa_sums(ps_trimmed_euk)[taxa_sums(ps_trimmed_euk)  <10] %>% length(),": ",round(taxa_sums(ps_trimmed_euk)[taxa_sums(ps_trimmed_euk)  <10] %>% length()/taxa_sums(ps_trimmed_euk)%>% length()*100,2)," %")

p1
```

### Effect of removing rare ASVs 

```{r, echo=FALSE, message=FALSE, warning=FALSE, collapse=TRUE}
# Filtering taxa -------------------------------------------------
ps_trimmed_euk_min = phyloseq::filter_taxa(ps_trimmed_euk %>% phyloseq_validate(), function(x) sum(x > 10) > (1), TRUE)

# Adding reads lost for sediment samples (processed in another script)
before = ps_trimmed_euk %>% sample_sums() %>% sum()
ps_trimmed_euk
after = ps_trimmed_euk_min %>% sample_sums() %>% sum()
ps_trimmed_euk_min
cat("\nPercent of lost reads after rare ASVs removal\n")
paste0("Lost reads: ",round(100-(after/before * 100),2)," %")
cat("\nPercent of lost ASVs after rare ASVs removal\n")
paste0("Lost ASVs: ",round(100-(ntaxa(ps_trimmed_euk_min)/ntaxa(ps_trimmed_euk) * 100),2)," %")
```


### Looking at sequencing depth per sample with rarefaction curves

```{r , message=FALSE, warning=FALSE, collapse=TRUE, include=FALSE}
ps_trimmed_euk_min = ps_trimmed_euk # Keeping rare taxa for now
ps_trimmed_euk_min <- prune_samples(sample_sums(ps_trimmed_euk_min)>100, ps_trimmed_euk_min)

dir.create(paste0(path_metabarcoding_data,"Cleaned_data"))
dir.create(paste0(path_metabarcoding_data,"Cleaned_data/Sequencing_depth"))

s = sample_sums(ps_trimmed_euk_min) %>% sort()
x = split(s, ceiling(seq_along(s)/10))
xx <- plyr::ldply(x, function(x){
    plyr::rename(ldply(x, function(z) {data.frame(read_count=z)}),c(".id" = ".id2"))
})
xx =xx %>% dplyr::group_by(.id) %>% dplyr::mutate(.id3 = max(read_count))
xx = xx[match(sample_names(ps_trimmed_euk_min),xx$.id2),]
ps_trimmed_euk_min@sam_data$read_count_gr = paste0("Group_",xx$.id," (<=",xx$.id3,")")
ps_trimmed_euk_min@sam_data$read_count_gr = factor(ps_trimmed_euk_min@sam_data$read_count_gr, levels = gtools::mixedsort(ps_trimmed_euk_min@sam_data$read_count_gr %>% unique()))
p <- ranacapa::ggrare(ps_trimmed_euk_min, step = 1000, color = "read_count_gr",  se = FALSE)
p <- p + facet_wrap(~read_count_gr,scales = "free_x") + theme_bw() + ylab("ASV richness") + xlab("Sequencing depth") + guides(color=guide_legend(title="Sequencing depth")) + theme(axis.text.x = element_text(angle = 45))
ggsave(plot = p, filename = "rarefaction_curves.pdf", path = paste0(path_metabarcoding_data,"Cleaned_data/Sequencing_depth/"), device = "pdf", width = 12, height = 8)
```

```{r echo=FALSE, fig.height= 8, fig.width=16}
ggplotly(p)
```

### Looking at read count per sample and removing those with less than X reads.

```{r echo=FALSE, message=FALSE, warning=FALSE}
df_read_count = ps_trimmed_euk_min %>% phyloseq::sample_sums() %>% sort() %>% as.data.frame(col.names = "Reads")
colnames(df_read_count) = "Read count"
df_read_count %>% datatable(class = 'cell-border stripe', caption = "Read count per sample after decontamination", options = list(pageLength = 10))

# Filtering samples -------------------------------------------------
ps_trimmed_euk_min_min = phyloseq::prune_samples(phyloseq::sample_sums(ps_trimmed_euk_min) >= 2000, ps_trimmed_euk_min)
```

### Saving results

```{r message=FALSE, warning=FALSE, code_folding=TRUE}
ps_trimmed_euk_min_min
saveRDS(ps_trimmed_euk_min_min,paste0(path_metabarcoding_data,"Cleaned_data/ps_clean.rds"))

# # Txt tables ----------------------------------------------------
ASV_table = ps_trimmed_euk_min_min %>% pstoveg_otu() %>% t() %>% as.data.frame() %>% rownames_to_column('feature-id') 
write.table(ASV_table,paste0(path_metabarcoding_data,'Cleaned_data/ASV_table.txt'), quote = F, row.names = F, sep = "\t")

Taxo_table = ps_trimmed_euk_min_min@tax_table@.Data %>% as.data.frame()
Taxo_table  = Taxo_table %>% rownames_to_column('feature-id')
write.table(Taxo_table,paste0(path_metabarcoding_data,"Cleaned_data/taxonomy_table.txt"), quote = F, row.names = F, sep = "\t")

Metadata = ps_trimmed_euk_min_min@sam_data %>% data.frame()
Metadata  = Metadata %>% rownames_to_column('sample-id')
write.table(Metadata,paste0(path_metabarcoding_data,"Cleaned_data/metadata_table.txt"), quote = F, row.names = F, sep = "\t")

writeXStringSet(ps_trimmed_euk_min_min@refseq,paste0(path_metabarcoding_data,"Cleaned_data/ref_seq.fasta"))
```



